Software for the analysis and visualization of deep mutational scanning data

### 2015-03-24 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex

\parttitle{Results} %if any
I describe a software package, \texttt{dms\_tools}, to infer the impacts of mutations from deep mutational scanning data using a likelihood-based treatment of the mutation counts. I show that \texttt{dms\_tools} yields more accurate inferences on simulated data than the widely used but statistically biased approach of calculating ratios of counts pre- and post-selection. Using \texttt{dms\_tools}, one can infer the preference of each site for each amino acid given a single selection pressure, or assess the extent to which these preferences change under different selection pressures. The preferences and their changes can be intuitively visualized with sequence-logo-style plots created using an extension to \texttt{weblogo}.

@@

\section*{Background}
Deep mutational scanning is a high-throughput experimental technique to assess the impacts of mutations on a protein-coding gene~\cite{fowler2014deep}. Figure \ref{fig:DeepMutationalScan} shows a schematic of deep mutational scanning. A gene is mutagenized, and the library of resulting variants is introduced into cells or viruses, which are then subjected to an experimental selection that enriches for functional variants and depletes non-functional ones. Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations. Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive,kitzman2015massively}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B. New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.

@@

The rest of this paper assumes that the sequencing depth is less than the number of unique molecules in the mutant library, such that the deep sequencing randomly subsamples the set of molecules. If this assumption is false (i.e. if the number of unique molecules is substantially less than the sequencing depth), then the accuracy of inferences about mutational effects will be fundamentally limited by this aspect of the experimental design. Properly done experiments should quantify the number of unique molecules in the library so that it is obvious whether this assumption holds. In the absence of such information, the analysis can be repeated using only a random fraction of the deep sequencing data to assess whether inferences are limited by sequencing depth or they underlying molecular diversity in the mutant library.

@@

\subsubsection*{Implementation.}
The program \texttt{dms\_inferprefs} in the \texttt{dms\_tools} package infers the preferences by using \texttt{pystan}~\cite{pystan-software:2014} to perform MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerrpre}, \ref{eq:pr_nrerrpost}, \ref{eq:pr_nrpre}, \ref{eq:pr_nrpost}, \ref{eq:pr_pir}, \ref{eq:pr_mur}, \ref{eq:pr_epsilonr}, and \ref{eq:pr_rhor}. The program runs four chains from different initial values, and checks for convergence by ensuring that the Gelman-Rubin statistic $\hat{R}$~\cite{gelman1992inference} is $<1.1$ and the effective sample size is $>100$; the number of MCMC iterations is increased until these convergence is achieved. The program \texttt{dms\_logoplot} in the \texttt{dms\_tools} package visualizes the posterior mean preferences via an extension to \texttt{weblogo}~\cite{crooks2004}. The program \texttt{dms\_merge} can be used to average preferences inferred from different experimental replicates that have individually been analyzed by \texttt{dms\_inferprefs}, and the program \texttt{dms\_correlate} can be used to compute the correlations among inferences from different replicates.

@@

Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} also illustrate use of \texttt{dms\_correlate} to assess the correlation between preferences inferred from different biological replicates~\cite{blainey2014points} of the experiment. The inclusion and analysis of such replicates provide the only sure way to fully assess the sources of noise associated with deep mutational scanning.

@@

\subsubsection*{Is the Bayesian inference worthwhile?}
The foregoing sections explain why the Bayesian inference of preferences implemented in \texttt{dms\_tools} is conceptually preferable to estimating mutational effects via direct calculation of enrichment ratios using Equation \ref{eq:enrichmentcounts}. However, do the practical benefits of this Bayesian inference justify its increased complexity? The simulations in the previous section show that the Bayesian inference is more accurate, but in the absence of background errors (Figure \ref{fig:PrefCorrelations}A) the magnitude of the improvement becomes negligible once the mutations counts per site $N \overline{\mu}$ start to exceed $\sim10^3$. When there is a need to correct for background errors (Figure \ref{fig:PrefCorrelations}B), meaningful benefits of the Bayesian inference over enrichment ratios extend to somewhat higher sequencing depths. Overall, it appears that Bayesian inference will always perform as well or better than enrichment ratios. But the tangible benefit becomes negligible at high sequencing depth. In that case, the user will have to decide if the increased computational runtime and complexity of the Bayesian inference is worth a small improvement in accuracy. Simpler enrichment-ratio calculations can be performed using the \texttt{--ratio\_estimation} option of \texttt{dms\_inferprefs} or using an alternative program such as \texttt{Enrich}~\cite{fowler2011enrich}. When applying enrichment-ratio calculations to data where some mutations have low counts, it is important to include pseudocounts (denoted by $\mathcal{P}$ in Equation \ref{eq:enrichmentcounts}) as a form of regularization to avoid estimating excessively high or low ratios at sites with limited counts.

@@

But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since we expect to alternative selections to have markedly different effects on only a small subset of mutations (typically, major constraints related to protein folding and stability will be relatively conserved across different selections on the same protein). Therefore, \texttt{dms\_tools} uses a different algorithm to infer the differential preferences under the two selections. This algorithm combines a prior that mildly favors differential preferences of zero with a likelihood-based analysis of the mutation counts to estimate posterior probabilities over the differential preferences.

@@

\subsubsection*{Likelihoods of observing specific mutational counts.}
Define vectors of the counts as $\mathbf{n_r^{\textbf{start}}}= \left(\cdots, n_{r,x}^{\textrm{start}}, \cdots\right)$ for the post-selection functional variants that are subjected to the further selections, and as $\mathbf{n_r^{s1}}= \left(\cdots, n_{r,x}^{s1}, \cdots\right)$ and $\mathbf{n_r^{s2}}= \left(\cdots, n_{r,x}^{s2}, \cdots\right)$ for selections $s1$ and $s2$. We again allow an error control, but now assume that the same control applies to all three libraries (since they are all sequenced after a selection), and define the counts for this control as $\mathbf{n_r^{\textbf{err}}}= \left(\cdots, n_{r,x}^{\textrm{err}}, \cdots\right)$; the true error frequencies are denoted by $\xi_{r,x}$. Define vectors of the frequencies, errors, control preferences, and differential preferences: $\boldsymbol{\mathbf{f_r^{\textbf{start}}}}= \left(\cdots, f_{r,x}^{\textrm{start}}, \cdots\right)$, $\boldsymbol{\mathbf{f_r^{s1}}}= \left(\cdots, f_{r,x}^{s1}, \cdots\right)$, $\boldsymbol{\mathbf{f_r^{s2}}}= \left(\cdots, f_{r,x}^{s2}, \cdots\right)$, $\boldsymbol{\mathbf{\xi_r}}= \left(\cdots, \xi_{r,x}, \cdots\right)$, $\boldsymbol{\mathbf{\pi^{s1}_r}}= \left(\cdots, \pi^{s1}_{r,x}, \cdots\right)$, and $\boldsymbol{\mathbf{\Delta\pi_r}}= \left(\cdots, \Delta\pi_{r,x}, \cdots\right)$. Equations \ref{eq:controlpi} and \ref{eq:deltapi} imply
$\boldsymbol{\mathbf{f_r^{s1}}}= \frac{\boldsymbol{\mathbf{f_r^{\textbf{starat}}}}\circ \boldsymbol{\mathbf{\pi_r^{s1}}}}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \boldsymbol{\mathbf{\pi_r^{s1}}}}$ and $\boldsymbol{\mathbf{f_r^{s2}}} = \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}$.

@@

\subsubsection*{Implementation.}
The program \texttt{dms\_inferdiffprefs} in the \texttt{dms\_tools} package infers the differential preferences by performing MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerr}, \ref{eq:pr_nrstart}, \ref{eq:pr_nrs1}, \ref{eq:pr_nrs2}, \ref{eq:pr_controlpir}, \ref{eq:pr_xir}, \ref{eq:pr_frstart}, and \ref{eq:pr_deltapi}. The MCMC is performed as described for the preferences, and characters can again be any of nucleotides, amino acids, or codons. The program \texttt{dms\_logoplot} visualizes the posterior mean differential preferences via an extension to \texttt{weblogo}~\cite{crooks2004}. In addition, \texttt{dms\_inferdiffprefs} creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$. These posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although be default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}.

@@

\section*{Data and code for figures in this paper.}
The data and computer code used to generate the figures are in the tagged version of the \texttt{dms\_tools} source code at \url{https://github.com/jbloom/dms_tools} in \texttt{examples} subdirectory. The \texttt{LaTex} source for this paper is in the \texttt{paper} subdirectory.

>>>>

\parttitle{Results} %if any
I describe a software package, \texttt{dms\_tools}, to infer the impacts of mutations from deep mutational scanning data using a likelihood-based treatment of the mutation counts. I show that \texttt{dms\_tools} yields more accurate inferences on simulated data than simply calculating ratios of counts pre- and post-selection. Using \texttt{dms\_tools}, one can infer the preference of each site for each amino acid given a single selection pressure, or assess the extent to which these preferences change under different selection pressures. The preferences and their changes can be intuitively visualized with sequence-logo-style plots created using an extension to \texttt{weblogo}.

@@

\section*{Background}
Deep mutational scanning is a high-throughput experimental technique to assess the impacts of mutations on a protein-coding gene~\cite{fowler2014deep}. Figure \ref{fig:DeepMutationalScan} shows a schematic of deep mutational scanning. A gene is mutagenized, and the library of resulting variants is introduced into cells or viruses, which are then subjected to an experimental selection that enriches for functional variants and depletes non-functional ones. Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations. Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive,kitzman2015massively}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B. New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,kitzman2015massively}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.

@@

The rest of this paper assumes that the sequencing depth is less than the number of unique molecules in the mutant library, such that the deep sequencing randomly subsamples the set of molecules. If this assumption is false (i.e. if the number of unique molecules is substantially less than the sequencing depth), then the accuracy of inferences about mutational effects will be fundamentally limited by this aspect of the experimental design. Properly done experiments should quantify the number of unique molecules in the library so that it is obvious whether this assumption holds. In the absence of such information, the analysis can be repeated using only a random fraction of the deep sequencing data to assess whether inferences are limited by sequencing depth or the underlying molecular diversity in the mutant library.

@@

\subsubsection*{Implementation.}
The program \texttt{dms\_inferprefs} in the \texttt{dms\_tools} package infers the preferences by using \texttt{pystan}~\cite{pystan-software:2014} to perform MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerrpre}, \ref{eq:pr_nrerrpost}, \ref{eq:pr_nrpre}, \ref{eq:pr_nrpost}, \ref{eq:pr_pir}, \ref{eq:pr_mur}, \ref{eq:pr_epsilonr}, and \ref{eq:pr_rhor}. The program runs four chains from different initial values, and checks for convergence by ensuring that the Gelman-Rubin statistic $\hat{R}$~\cite{gelman1992inference} is $<1.1$ and the effective sample size is $>100$; the number of MCMC iterations is increased until convergence is achieved. The program \texttt{dms\_logoplot} in the \texttt{dms\_tools} package visualizes the posterior mean preferences via an extension to \texttt{weblogo}~\cite{crooks2004}. The program \texttt{dms\_merge} can be used to average preferences inferred from different experimental replicates that have individually been analyzed by \texttt{dms\_inferprefs}, and the program \texttt{dms\_correlate} can be used to compute the correlations among inferences from different replicates.

@@

Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} also illustrate use of \texttt{dms\_correlate} to assess the correlation between preferences inferred from different biological replicates~\cite{blainey2014points} of the experiment. The inclusion and analysis of such replicates is the only sure way to fully assess the sources of noise associated with deep mutational scanning.

@@

\subsubsection*{Is the Bayesian inference worthwhile?}
The foregoing sections explain why the Bayesian inference of preferences implemented in \texttt{dms\_tools} is conceptually preferable to estimating mutational effects via direct ratio estimation using Equation \ref{eq:enrichmentcounts}. However, do the practical benefits of this Bayesian inference justify its increased complexity? The simulations in the previous section show that the Bayesian inference is more accurate, but in the absence of background errors (Figure \ref{fig:PrefCorrelations}A) the magnitude of the improvement becomes negligible once the mutation counts per site $N \overline{\mu}$ start to exceed $\sim10^3$. When there is a need to correct for background errors (Figure \ref{fig:PrefCorrelations}B), meaningful benefits of the Bayesian inference over enrichment ratios extend to somewhat higher sequencing depths. Overall, it appears that Bayesian inference will always perform as well or better than ratio estimation, but that the tangible benefit becomes negligible at high sequencing depth. In that case, the user will have to decide if the increased computational runtime and complexity of the Bayesian inference is worth a small improvement in accuracy. Simpler ratio estimation can be performed using the \texttt{--ratio\_estimation} option of \texttt{dms\_inferprefs} or using an alternative program such as \texttt{Enrich}~\cite{fowler2011enrich}. When applying ratio estimation to data where some mutations have low counts, it is important to include pseudocounts (denoted by $\mathcal{P}$ in Equation \ref{eq:enrichmentcounts}) as a form of regularization to avoid estimating excessively high or low preferences at sites with limited counts.

@@

But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since we expect alternative selections to have markedly different effects on only a small subset of mutations (typically, major constraints related to protein folding and stability will be conserved across different selections on the same protein). Therefore, \texttt{dms\_tools} uses a different algorithm to infer the differential preferences under the two selections. This algorithm combines a prior that mildly favors differential preferences of zero with a likelihood-based analysis of the mutation counts to estimate posterior probabilities over the differential preferences.

@@

\subsubsection*{Likelihoods of observing specific mutational counts.}
Define vectors of the counts as $\mathbf{n_r^{\textbf{start}}}= \left(\cdots, n_{r,x}^{\textrm{start}}, \cdots\right)$ for the post-selection functional variants that are subjected to the further selections, and as $\mathbf{n_r^{s1}}= \left(\cdots, n_{r,x}^{s1}, \cdots\right)$ and $\mathbf{n_r^{s2}}= \left(\cdots, n_{r,x}^{s2}, \cdots\right)$ for selections $s1$ and $s2$. We again allow an error control, but now assume that the same control applies to all three libraries (since they are all sequenced after a selection), and define the counts for this control as $\mathbf{n_r^{\textbf{err}}}= \left(\cdots, n_{r,x}^{\textrm{err}}, \cdots\right)$; the true error frequencies are denoted by $\xi_{r,x}$. Define vectors of the frequencies, errors, control preferences, and differential preferences: $\boldsymbol{\mathbf{f_r^{\textbf{start}}}}= \left(\cdots, f_{r,x}^{\textrm{start}}, \cdots\right)$, $\boldsymbol{\mathbf{f_r^{s1}}}= \left(\cdots, f_{r,x}^{s1}, \cdots\right)$, $\boldsymbol{\mathbf{f_r^{s2}}}= \left(\cdots, f_{r,x}^{s2}, \cdots\right)$, $\boldsymbol{\mathbf{\xi_r}}= \left(\cdots, \xi_{r,x}, \cdots\right)$, $\boldsymbol{\mathbf{\pi^{s1}_r}}= \left(\cdots, \pi^{s1}_{r,x}, \cdots\right)$, and $\boldsymbol{\mathbf{\Delta\pi_r}}= \left(\cdots, \Delta\pi_{r,x}, \cdots\right)$. Equations \ref{eq:controlpi} and \ref{eq:deltapi} imply
$\boldsymbol{\mathbf{f_r^{s1}}}= \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \boldsymbol{\mathbf{\pi_r^{s1}}}}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \boldsymbol{\mathbf{\pi_r^{s1}}}}$ and $\boldsymbol{\mathbf{f_r^{s2}}} = \frac{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\circ \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}{\boldsymbol{\mathbf{f_r^{\textbf{start}}}}\cdot \left(\boldsymbol{\mathbf{\pi_r^{s1}}} + \boldsymbol{\mathbf{\Delta\pi_r}}\right)}$.

@@

\subsubsection*{Implementation.}
The program \texttt{dms\_inferdiffprefs} in the \texttt{dms\_tools} package infers the differential preferences by performing MCMC over the posterior defined by the product of the likelihoods and priors in Equations \ref{eq:pr_nrerr}, \ref{eq:pr_nrstart}, \ref{eq:pr_nrs1}, \ref{eq:pr_nrs2}, \ref{eq:pr_controlpir}, \ref{eq:pr_xir}, \ref{eq:pr_frstart}, and \ref{eq:pr_deltapi}. The MCMC is performed as described for the preferences, and characters can again be any of nucleotides, amino acids, or codons. The program \texttt{dms\_logoplot} visualizes the posterior mean differential preferences via an extension to \texttt{weblogo}~\cite{crooks2004}. In addition, \texttt{dms\_inferdiffprefs} creates text files that give the posterior probability that $\Delta\pi_{r,x} > 0$ or $< 0$. These posterior probabilities are \emph{not} corrected to account for the fact that multiple sites are typically being examined, although by default the inferences are made using the regularizing prior in Equation \ref{eq:pr_deltapi}.

@@

\section*{Data and code for figures in this paper.}
The data and computer code used to generate the figures are in version 1.01 of the \texttt{dms\_tools} source code (which is tagged on \texttt{Github} at \url{https://github.com/jbloom/dms_tools/tree/1.0.1}) in the \texttt{examples} subdirectory. The \texttt{LaTex} source for this paper is in the \texttt{paper} subdirectory.
